# Modeling trends

## Resources

[Slides](slides/04_trends/index.html)

## Activity

### Testing for a difference in time-varying pattern across two groups

:::{.try}

For the (made up) data below, you have longitudinal data for two groups, A and B, both of which show a quadratic trend over time. The data are in the R binary file [quadratic.rds](data/quadratic.rds){target="_download"}.

```{r load-data, echo=FALSE}
library("tidyverse")

qdata <- readRDS("data/quadratic.rds")
```

```{r plot-data, echo=FALSE}
ggplot(qdata, aes(time, response)) +
  geom_point(aes(color = group, shape=group))
```

Once you have downloaded the file and put it in your working directory, you can load the data using:

```{r load-data-show, eval=FALSE}
library("tidyverse")

qdata <- readRDS("quadratic.rds")
```

**Your task**: Using a single model, fit quadradic functions for both of the groups, with orthogonal polynomials.

If you need help adding polynomial predictors to your data frame, check out the function [`poly_add_columns()`](https://rdrr.io/github/tjmahr/polypoly/man/poly_add_columns.html) from the **`polypoly`** R package (install it using `install.packages("polypoly")`).

Note: You'll probably want to deviation or sum code your categorical predictor, for all the reasons mentioned in the last section on interactions.

*Need help building your data? Here's the first part of the solution.*

```{r build-data, webex.hide="solution for creating the tibble"}
library("polypoly")

qdata2 <- qdata %>%
  mutate(grp = if_else(group == "B", 1/2, -1/2)) %>%
  poly_add_columns(time, degree = 2, prefix = "t")
```

Now fit the model, then compare your result with the solution below.

```{r fit-model, webex.hide="solution for fitting the model"}
mod <- lm(response ~ (t1 + t2) * grp, data = qdata2)

summary(mod)
```

**Extra: if you wanted to simultaneously test whether the time-varying (linear, quadradic) components differed across the groups (a single p-value), how would you do this?**

```{r mod-compare, webex.hide="solution"}
mod2 <- update(mod, . ~ . -t1:grp -t2:grp)

anova(mod, mod2)
```

**Advanced: plot the data along with the model predictions for the two groups.**

Hint: `fitted()`

```{r show-plot-code, webex.hide="hint: see the code that made the graph above", ref.label="plot-data", eval=FALSE}
```

```{r plot-curves, webex.hide="solution"}
qdata3 <- qdata2 %>%
  mutate(fitted = fitted(mod))

ggplot(qdata, aes(time, response)) +
  geom_point(aes(color = group, shape=group)) +
  geom_line(data = qdata3, aes(y = fitted, color = group))
```

:::
